---
title: "Topic 2.3"
author: "Roi Naveiro"
date: "Academic Year 2026-2027"
output: 
   xaringan::moon_reader:
    css: "ds_slides.css"
    lib_dir: libs
    seal: false
    includes:
      after_body: insert-logo.html
    nature: 
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false
      slideNumberFormat: "%current%"  
---



```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=7, fig.height=5, fig.align='center') 
```

class: center, middle, inverse, title-slide

<div class="title-logo"></div>

<br>

# Data Analysis and Information Extraction

## Topic 2 - Exploratory Data Analysis

### 2.3 Exploratory Data Analysis

<br>
.pull-left[
### Roi Naveiro
#### Academic Year 2026-2027
]
---

## Exploratory Data Analysis

* We will learn how to explore data systematically...

* ...using the tools we have learned (visualization and transformation)

* Exploratory Data Analysis is a cycle consisting of

  1. Generating questions about the data
  2. Answering them by visualizing, transforming and modeling
  3. Using what we have learned to refine questions or raise new ones

---

## Exploratory Data Analysis

* **Goal**: to develop an understanding of the data

* It is useful to do this through questions

  - They allow us to focus on one part of the data
  - They allow us to decide which plots, models and transformations to use
  
* A **creative** process

---

## Exploratory Data Analysis

* A **creative** process: there are no rules for determining which questions are the most useful.

* Two types of questions that are (almost always) interesting:

  - What type of variation is present in my variables?
  - What type of co-variation appears **between** my variables?
  
---
## A bit of jargon

* **Variable**: a property that can be measured
* **Value**: the state of the variable when it is measured. It can change if the measurement is repeated!
* **Observation**: a set of measurements taken under similar conditions (e.g. on the same subject). Several values, each one associated with a different variable
* **Tabular Data**: a set of values, each one associated with a variable and an observation. Variables in columns, observations in rows.


---
class: center, middle, inverse

## EDA: Variation

---
## Variation

*Variation is the tendency of the values of a variable to change from measurement to measurement*

This variation can be associated with different phenomena:

- **Measurement error** when measuring a constant quantity (the speed of light)
- Measurement of the same variable on **different subjects**
- Measurement of the same variable, on the same subject, at **different times**
- ...

The pattern of variation gives interesting information...

... which is distilled by visualizing the **distribution of the values of the variable**

---

## Variation: categorical variable

Let us recall how to visualize the distribution of a **categorical variable**

```{r}
library(ggplot2)
# We plot the number of gears in the mtcars data
ggplot(data=mtcars, aes(x=gear)) + 
  geom_bar() + 
  labs(x = "Number of gears")
```

---
## Variation: categorical variable

The height of the bars shows the number of observations. 
How would you obtain this number?

--

```{r}
library(tidyverse)
mtcars %>% group_by(gear)  %>% summarise(n())
```

---
## Variation: continuous variable

We had also seen how to visualize the distribution of continuous variables
```{r}
ggplot(data=iris, aes(x=Petal.Length)) + 
  geom_histogram(bins = 20) + 
  labs(x = "Petal Length (cm)")
```



---
## Variation

What should we look for in variation plots? What questions should we ask?

* Typical values

* Outliers

---
## Variation: typical values

- Which values are the most common?

- Which are the least common?

- Is there any unusual pattern?

---
## Example: petal length

```{r}
ggplot(data=iris, aes(x=Petal.Length)) + 
  geom_histogram() + 
  labs(x = "Petal Length (cm)")
```

---
## Example: petal length

*Clusters* of similar values suggest that there are subgroups in the data. To understand them, it is worth asking

* What do the observations of each subgroup have in common?

* What distinguishes the observations of different subgroups?

* How can we explain the subgroups?

--

* There is a group of flowers with short petals (<2cm) and another with long petals (>2cm)

* Does the species variable explain this separation?

---
## Variation: continuous variable

* If we want to visualize several histograms in the same plot, we can use colors

* Sometimes it is clearer to use `geom_freqpoly()`

* The same as a histogram, but drawn with polygonal lines

---
## Variation: continuous variable

```{r}
ggplot(data=iris, aes(x=Petal.Length, fill=Species)) + 
  geom_histogram(bins = 20) + 
  labs(x = "Petal Length (cm)")
```


---
## Variation: continuous variable

```{r}
ggplot(data=iris, aes(x=Petal.Length, color=Species)) + 
  geom_freqpoly() + 
  labs(x = "Petal Length (cm)")
```

---
## Variation: outliers

* Finding **outliers** is as important as studying typical values

* Outliers:

  - Sometimes, measurement errors
  - Other times, important discoveries!
  
---
## Example: diamond width

Let us study the distribution of the width of the diamonds in the `diamonds` dataset (more information in `?diamonds`)

```{r}
ggplot(diamonds, aes(x=y)) +
  geom_histogram(binwidth = 0.5)
```

---
## Example: diamond width

Apparently nothing strange, but...

```{r}
ggplot(diamonds, aes(x=y)) +
  geom_histogram(binwidth = 0.5) + 
  coord_cartesian( ylim = c(0,50) )
```

---
## Example: diamond width

It seems that there are outliers around 0, 30 and 60.

Let us extract them with `filter`
--

```{r}
diamonds %>%
  filter(y < 3 | y > 20) 
```

---
## Example: diamond width

- It seems that there are diamonds of size 0. Incorrect data.

- There are very large diamonds, but not too expensive...

---
## Variation: outliers

- It is usually useful to repeat the analysis with and without outliers, in order to measure their effect

- If the effect is minimal and we do not know their origin, replace them with `NA`

- Otherwise, remove them only in a **justified** way and report it

---
class: center, middle, inverse

## EDA: Missing values

---
## Outliers

When we want to remove outliers, there are two options:

- Remove the whole row (a lot of information is lost)

```{r}
diamonds2 <- diamonds %>%
  filter(y>=3 & y<=20)
```

- Replace outliers with `NA`

```{r}
diamonds2 <- diamonds %>%
  mutate(y = ifelse(y<3 | y>20, NA, y))
```

- **NOTE**: the first argument of `ifelse()` is logical, the second is the value if `TRUE` and the third, the value if `FALSE`

---
## Missing values

Missing values are indicated with a warning
```{r, warning=T}
ggplot(data=diamonds2, mapping = aes(x=x, y=y))+
  geom_point()
```

---
## Missing values

It can be removed with `na.rm = TRUE`
```{r, warning=T}
ggplot(data=diamonds2, mapping = aes(x=x, y=y))+
  geom_point(na.rm = T)
```


---
class: center, middle, inverse

## EDA: Covariation

---

## Covariation

- The tendency of the values of two or more variables to change together

- The way to visualize covariation depends on the type of variables (as we saw)

  * Categorical - Continuous
  * Categorical - Categorical
  * Continuous - Continuous
  
  
---

## Covariation: Categorical - Continuous

```{r}
ggplot(data=diamonds, aes(x=price, color=cut)) + 
  geom_freqpoly()
```

---

## Covariation: Categorical - Continuous

`..density..` plots density on the y axis, which is the count standardized so that the area of each polygon is 1
```{r}
ggplot(data=diamonds, aes(x=price, y=..density.., color=cut)) + 
  geom_freqpoly()
```

---

## Covariation: Categorical - Continuous

What other way is there to visualize covariation between categorical and continuous variables?
--

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) + 
  geom_boxplot()
```

---

## Covariation: Categorical - Continuous

Sometimes, it is worth reordering...

```{r}
ggplot(data = mpg, aes(x = reorder(class, hwy, FUN = median), y = hwy)) + #<<
  geom_boxplot()
```

---

## Covariation: Categorical - Categorical

We had seen bar charts. Another way:

```{r}
ggplot(data = diamonds, aes(x = cut, y = color)) + 
  geom_count() #<<
```
---

## Covariation: Categorical - Categorical

How would you do this with `dplyr`?
--

```{r}
diamonds %>%
  group_by(color, cut) %>%
  summarise(n = n())
```
---

## Covariation: Categorical - Categorical

This also works

```{r}
diamonds %>%
  count(color, cut) 
```

---
## Covariation: Categorical - Categorical

With this

```{r, fig.width=6, fig.height=4}
diamonds %>%
  count(color, cut) %>%
  ggplot(aes(x = color, y = cut, fill = n))+ 
  geom_tile() #<<
```

---
## Covariation: Continuous - Continuous

- We saw how to do it with `geom_point()`

- In very large datasets, the points overlap

- Idea? Use `alpha`

---
## Covariation: Continuous - Continuous

```{r}
ggplot(data=diamonds, aes(x = carat, y = price)) + 
  geom_point(alpha = 0.01) #<<
```

---
## Covariation: Continuous - Continuous

Better

```{r}
ggplot(data=diamonds, aes(x = carat, y = price)) + 
  geom_bin2d() #<<
```


---
## Covariation: Continuous - Continuous

Better

```{r}
# install.packages("hexbin")
ggplot(data=diamonds, aes(x = carat, y = price)) + 
  geom_hex() #<<
```

---
## Covariation: Continuous - Continuous

Another option

```{r}
ggplot(data=diamonds, aes(x = carat, y = price, )) + 
  geom_boxplot(aes(group = cut_width(carat, 0.1)) ) #<<
```


---
## Covariation: Continuous - Continuous

Another option (the same number of points in each bin)

```{r}
ggplot(data=diamonds, aes(x = carat, y = price, )) + 
  geom_boxplot(aes(group = cut_number(carat, 20)) ) #<<
```

---
class: center, middle, inverse

## EDA: Patterns and models

---

## Patterns and models

Patterns reveal clues about relationships between variables. When we find one, it is worth asking

- Could it be due to chance?

- How is the relationship it implies described?

- How strong is it?

- What other variables affect this relationship?

- Does the relationship change if we look at subgroups of the data?


**Models** are used to extract patterns from the data...

... e.g. they can be used to remove the effect of one variable and study the remaining effect.

---

## Bibliography

* [R for Data Science](https://r4ds.had.co.nz/), Wickham and Grolemund (2016)
