---
title: "Lab 1 - Programming in R"
output:
  html_document:
    css: ./ds_labs.css
    toc: yes
    toc_float: yes
    fig_caption: no
    seal : False
    includes:
      after_body: insert-logo.html
  pdf_document:
    toc: yes
---

<div class="title-logo"></div>

```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=8, fig.height=6) 
```


## Exercise 1

Write a function that returns $k$ numbers, where each number is
the sum of $n$ numbers drawn at random between 1 and 100 (with replacement), divided by $n$.

Plot a histogram of the results for $k=10000$ with $n=1$, $n=2$, $n=10$ and $n=100$.
Hint: you can use the `hist()` function.

What do you observe?


## Exercise 2

Use the help page of the `replicate()` function to learn how it works.
Solve Exercise 1 without using a loop (if you used one).


## Exercise 3

R's `mtcars` dataset contains information extracted from the *1974 Motor Trend US magazine* about 10 aspects of design and performance for 32 vehicles. 
Using this dataset, compute 

**Question 1.** The difference between the average fuel consumption of automatic-transmission vehicles and that of manual-transmission ones.

**Question 2.** The same difference, but only for vehicles weighing more than 3000 lbs.

Hint: the `?mtcars` help page may be useful

## Exercise 4

Using a `for` loop and an `if`, add up the even numbers between 1 and 100 (both included).

## Exercise 5

Compute the expected gain of the roulette strategy that consists of always betting 10 euros on red.

To do so:

  - Create a function called `spin()` that returns a number drawn at random between 0 and 36 (both included).
  - Create another function called `compute_gain()` with an argument `q` for the amount of money bet, with a default value of 10. This function has to generate one spin of the roulette wheel and, depending on the outcome, compute the gain and return it. Remember that
  
    * The numbers (32, 19, 21, 25, 34, 27, 36, 30, 23, 5, 16, 1, 14, 9, 18, 7, 12, 3) are red.
    * The numbers (15, 4, 2, 17, 6, 13, 11, 8, 10, 24, 33, 20, 31, 22, 29, 28, 35, 26) are black.
    * Zero has no colour.
    
  - Finally, create a loop that simulates `K = 10000` rounds of roulette, storing the gain generated in each one, thus building a vector of length $K$. Call this vector `winnings`. Compute its mean. What do you observe?
  
## Exercise 6

Use the following code to plot the cumulative winnings against the round number. You will need to have created the `winnings` vector from the previous exercise and to have defined the parameter `K` (number of rounds). Draw a conclusion from the plot.

```{r, eval=F}
library(ggplot2)

results = data.frame(Round = 1:K, Cumulative_Winnings = cumsum(winnings) )

p <- ggplot(results, aes(x=Round, y=Cumulative_Winnings)) +  
  geom_line() + theme_minimal()

p
```


## Exercise 7

Repeat the two previous exercises, but this time evaluating any other strategy you can think of. What conclusion do you draw?

## Exercise 8

Write R code to estimate the value of $\pi$ using the Monte Carlo method. The method consists of generating random points in the unit square and counting how many of them fall inside a quarter circle inscribed in that square.

1. Generate $N$ random points inside a square of side 1, that is, the points $(x, y)$ must have coordinates $x$ and $y$ between 0 and 1. You can use the `runif()` function for this.

2. Compute the number of points that fall inside the quarter circle. 

3. Use the number computed in the previous step and the total number of points generated to estimate the value of $\pi$.

What do you observe as you increase the number of points generated?





