---
title: "Lab 2.1 - Data Visualization"
output:
  html_document:
    css: ./ds_labs.css
    toc: yes
    toc_float: yes
    fig_caption: no
    seal : False
    includes:
      after_body: insert-logo.html
  pdf_document:
    toc: yes
---


<div class="title-logo"></div>

```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
```


## Exercise 1

Using the data from the `gapminder` package:

* Filter the observations corresponding to the years 1957, 1967, 1977, 1987, 1997, 2007

* Plot the relationship between GDP per capita and life expectancy by continent and by year
in different plots using **faceting**. The size of the points must be proportional to the population.
The color must reflect the continent. 

* What conclusions do you draw?

## Exercise 2

Pollution due to plastic consumption is a global problem that negatively affects the health of many ecosystems.
[Our World in Data](https://ourworldindata.org/plastic-pollution) contains a great deal of data on the evolution of plastic consumption both globally and by country. For this exercise we focus on the 2010 data.

The data are in the data folder of the repository. You can load them using

```{r load-data, message=FALSE, eval=TRUE}
library(tidyverse)
plastic <- read_csv("data/plastic-waste.csv")
```

The variables are the following

- `code`: Country code (3 letters)
- `entity`: Full name of the country
- `continent`: Name of the continent
- `year`: Year
- `gdp_per_cap`: GDP per capita (2011) in international dollars
- `plastic_waste_per_cap`: Amount of plastic waste per capita in kg/day
- `mismanaged_plastic_waste_per_cap`: Amount of mismanaged plastic waste in kg/day
- `mismanaged_plastic_waste`: Tonnes of mismanaged plastic waste
- `coastal_pop`: Number of individuals living on or near the coast
- `total_pop`: Total population

Visualize the distribution of the amount of plastic waste per capita (in 2010).

Are there any outliers? If so, determine which countries are outliers. Try to establish whether or not they are erroneous data (by researching on the internet).

Note: For the remaining exercises, filter out the outliers if there are any.


## Exercise 3

Visualize the distribution of the amount of plastic waste per capita by continent, using a different histogram for each continent.


## Exercise 4

Repeat the previous plot using density, including all the continents in the same plot and indicating them with different colors.


## Exercise 5

Improve the previous plot by filling each curve with a color and adding transparency.


## Exercise 6

Produce an alternative visualization to the previous one that allows the distribution of the amount of plastic waste per capita of different continents to be compared.


## Exercise 7

Find out what [violin plots](http://ggplot2.tidyverse.org/reference/geom_violin.html) are.
Build the violin plot of the previous data. What do these plots reveal that a boxplot does not?

Do you observe any pattern that distinguishes the distribution of the amount of plastic waste of developed and developing continents?


## Exercise 8

Visualize the relationship between the amount of plastic waste per capita and GDP per capita. Color each continent with a different color and add a **single smoothing** that shows the overall trend. In addition, use the viridis color scale (look up on the internet how to use it).
How would you describe this relationship? 


## Exercise 9

Repeat the previous plot but including only the data for Europe (obviously without using colors).
What trend do you observe? Which countries consume less plastic than this trend implies?


## Exercise 10

Visualize the relationship between plastic waste per capita and mismanaged waste per capita.
Color each continent with a different color.
Does the association between these two variables change across continents?

