---
title: "Lab 3.3 - Strings, Factors and Dates"
output:
  html_document:
    css: ./ds_labs.css
    toc: yes
    toc_float: yes
    fig_caption: no
    seal : False
    includes:
      after_body: insert-logo.html
  pdf_document:
    toc: yes
---


<div class="title-logo"></div>

```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=8, fig.height=6) 
```


```{r}
library(tidyverse)
library(nycflights13)
library(lubridate)
library(tidyquant)
```


## Exercise 1

In the `flights` dataset the variables `dep_time`, `arr_time`, `sched_dep_time`
and `sched_arr_time` contain the times in three- or four-digit format.
Thus, 830 is half past eight in the morning and 2030 is half past eight in the evening.
One way to extract the hour and the minutes from these times is the following:

```{r}
time <- 2030
hour <- time%/%100
minutes <- time%%100
print(hour)
print(minutes)
```

Understand and describe how these lines of code work. Create a dataframe called `flights_dt` in which the variables `dep_time`, `arr_time`, `sched_dep_time`
and `sched_arr_time` contain the exact dates they refer to, in `datetime` format.

### Solution


## Exercise 2

Visualize the distribution of flights by the **day of the week** on which they take off.
What conclusion do you draw?

### Solution


## Exercise 3

Visualize the distribution of flights by the **minute** at which they take off and by the **minute**
of the scheduled departure time.
What pattern appears? How would you explain it?

### Solution


## Exercise 4

Create a new variable corresponding to the difference between `arr_time` and
`dep_time`. 
Visualize the covariation between this variable and `air_time`.
What do you observe? What is happening?

### Solution

## Exercise 5

Repeat the previous visualization, but using a different color for each time 
zone of the destination airport. What do you observe?

**Hint**: the time zone of the different airports is in the 
`airports` dataset and is called `tzone`.


### Solution


## Exercise 6

The `tidyquant` library can be used to import countless economic datasets into R. In particular, it can be used to import data on the price
of different stocks.

Extracting the price history of AMZN since the year 2000 is very simple:

```{r}
prices  <- tq_get("AMZN", get = "stock.prices", from = " 2000-01-01")
prices 
```

Create a visualization that allows you to study the variation of the AMZN 
opening price over time. On what dates are the maximum and minimum 
opening prices reached?

At the start of the pandemic, because of the mandatory lockdowns imposed by
many countries, market expectations for Amazon soared.
Imagine you had bought Amazon shares on 2 March 2020 for
a value of 5000 euros. Compute the net gains you would have obtained if you had sold
those shares one year later. (Use the opening prices).

### Solution


## Exercise 7

When using regular expressions, the `+` sign indicates repetition.
Thus, if we want to detect a pattern that repeats an indefinite number of times,
for example, detect all the Cs in CCCCCCCMMD, we proceed as follows.

```{r}
str_view("CCCCCCCMMD", pattern = "C+")
```
In addition, the `str_extract` function extracts the detected patterns. For example
```{r}
str_extract("CCCCCCCMMD", pattern = "C+")
```

Using regular expressions, extract the ages from the following sentences.
The result must be a vector containing the numbers 57, 36, 23 and 1.
```{r}
x1 <- "The patient is 57 years old"
x2 <- "My sister's age is 36 years"
x3 <- "At 23 years of age, she wore a size 36 dress"
x4 <- "He is only 1 year old!"

texts <- c(x1,x2,x3,x4)
```


### Solution

## Exercise 8
Using the `gss_cat` data from tidyverse, visualize the distribution of the `rincome` variable.
To improve the visualization, recode the levels of this variable so that they are
`Unknown`, `<$15000` and `>$15000`.

### Solution

## Exercise 9

The data in `amazon_reviews.csv` contain reviews and ratings that certain users have given to a series of Amazon products. The variables included are:

* `review`: text of the review
* `rate`: rating (from 1 to 5)

Create a new variable called `n_loves` corresponding to the number of times the word "love" or "Love" appears in each review. Compute the mean of `n_loves` for each value of `rate`. Present the results in a single tibble and draw a conclusion.

### Solution

