---
title: "Lab 4 - Models"
output:
  html_document:
    css: ./ds_labs.css
    toc: yes
    toc_float: yes
    fig_caption: no
    seal : False
    includes:
      after_body: insert-logo.html
  pdf_document:
    toc: yes
---


<div class="title-logo"></div>

```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=8, fig.height=6) 
```


```{r}
library(tidyverse)
library(modelr)
```


## Exercise 1

Using the `diamonds` dataset, visualize the relationship between the following
pairs of variables:

* `cut` and `price`
* `color` and `price`
* `clarity` and `price`

Do you find an unusual pattern?

**Hint**: the lowest-quality color for a diamond is J (slightly yellow), the worst clarity is I1 and the worst cut is Fair.


### Solution

```{r}
ggplot(diamonds, aes(cut, price)) + geom_boxplot()
ggplot(diamonds, aes(color, price)) + geom_boxplot()
ggplot(diamonds, aes(clarity, price)) + geom_boxplot()
```
Low-quality diamonds are more expensive

## Exercise 2

Fit a linear model where the predictor variable is `carat` (the weight of the diamonds) and the dependent variable is `price`. Obtain the value of the coefficients and interpret them. What conclusion do you draw?


### Solution

```{r}
mod_diamond <- lm(price ~ carat, data = diamonds)
summary(mod_diamond)
```

## Exercise 3

Visualize the predictions of the previous model, as well as the residuals.
Draw conclusions.


### Solution


```{r}
grid <- diamonds %>% 
  data_grid(carat = seq_range(carat, 40)) %>% 
  add_predictions(mod_diamond, "price") 


diamonds %>% add_residuals(mod_diamond, "resid")


ggplot(diamonds, aes(carat, price)) + 
  geom_point() + 
  geom_line(data = grid, colour = "red")
```

## Exercise 4

Create a new dataset called `diamonds2` in which you remove all
diamonds with carat greater than 2.5. Add two variables to this dataset
called `lcarat` and `lprice` corresponding respectively to the base-two 
logarithm of `carat` and `price`.

### Solution

```{r}
diamonds2 <- diamonds %>% 
  filter(carat <= 2.5) %>% 
  mutate(lprice = log2(price), lcarat = log2(carat))
```



## Exercise 5

Using the data created in the previous exercise, fit a linear model where the predictor variable is `lcarat` and the dependent variable is `lprice`. Obtain the value of the coefficients and interpret them. According to this model, what happens to the price of a diamond if I double its `carat`?

### Solution

```{r}
mod_diamond <- lm(lprice ~ lcarat, data = diamonds2)
coefficients(mod_diamond)
```

$$
\log_2 y = \beta_0 + \beta_1 \log_2 x
$$

$$
y = 2^{\beta_0} x^{\beta_1}
$$

$$
y = 2^{\beta_0} (2x)^{\beta_1} = 2^{\beta_0} (x)^{\beta_1} 2^{\beta_1}
$$


## Exercise 6

Visualize the predictions of the previous model, as well as the residuals.
Draw conclusions. Does this model fit the data better than the previous one?

### Solution

```{r}
grid <- diamonds2 %>% data_grid(lcarat) %>% 
  add_predictions(mod_diamond, "lpred")

ggplot(diamonds2, aes(lcarat)) + 
  geom_point( aes(y=lprice) ) + 
  geom_line(data = grid, aes(y=lpred), colour = "red", size = 1)
```


```{r}
grid <- diamonds2 %>% 
  data_grid(carat) %>% 
  mutate(lcarat = log2(carat)) %>% 
  add_predictions(mod_diamond, "lprice") %>% 
  mutate(price = 2 ^ lprice)


ggplot(diamonds2, aes(carat, price)) + 
  geom_point() + 
  geom_point(data = grid, colour = "red", size = 1)

```

## Exercise 7

Using the `diamonds2` dataset, visualize the relationship between the following variables and the residual computed in the previous exercise.
* `cut` 
* `color`
* `clarity`

What conclusion do you draw?

### Solution

```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond, "lresid")

ggplot(diamonds2, aes(lcarat, lresid)) + 
  geom_point()
```

```{r}
ggplot(diamonds2, aes(cut, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(color, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(clarity, lresid)) + geom_boxplot()
```



## Exercise 8

Explore the `sim3`  data from `modelr`. What type are the variables `x1` and `x2`? And `y`?
Fit a linear model with response variable `y` and predictors `x1` and `x2`  and visualize
its predictions.

Interpret the coefficients of this model.


### Solution

```{r}
ggplot(sim3, aes(x1, y)) + 
  geom_point(aes(colour = x2))
```

```{r}
mod1 <- lm(y ~ x1 + x2, data = sim3)

coef(mod1)

grid <- sim3 %>% 
  data_grid(x1, x2) %>% 
  add_predictions(mod1)

ggplot(sim3, aes(x1, y, colour = x2)) + 
  geom_point() + 
  geom_line(data = grid, aes(y = pred)) 
```

$$
y = \beta_0 + \beta_1x_1 + \beta_2 x_{2b} + \beta_3 x_{2c}+ \beta_4 x_{2d}
$$

$$
y = \beta_0 + \beta_1 x + \beta_4 
$$

$$
y = \beta_0 + \beta_1 x
$$


$$
y = \beta_0 + \beta_1 (x+1) + \beta_3 = y + \beta_1
$$

## Exercise 9

With the previous data, fit the following model

$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2
$$

in R this model corresponds to the formula `y ~ x1 * x2`.
Visualize the predictions of the model, and interpret its coefficients.

### Solution

```{r}
mod2 <- lm(y ~ x1 * x2, data = sim3)
grid <- sim3 %>% 
  data_grid(x1, x2) %>% 
  gather_predictions(mod2)

ggplot(sim3, aes(x1, y, colour = x2)) + 
  geom_point() + 
  geom_line(data = grid, aes(y = pred)) 
```

## Exercise 10

For this exercise you need to load the `infer` package

```{r load-packages, warning = FALSE, message = FALSE}
library(infer)
```

### Data 


The General Social Survey (GSS) collects data on contemporary American society in order to understand and explain trends and constants in attitudes, behaviors and attributes. Hundreds of trends have been monitored since 1972. 

The GSS contains a standard core of demographic, behavioral and attitudinal questions, plus topics of special interest. Among the topics covered are civil liberties, crime and violence, intergroup tolerance, morality, national spending priorities, psychological well-being, social mobility and stress and traumatic events.

We will analyze GSS data from 2018 to understand how much time American adults spend on email.

```{r load-data, message = FALSE}
gss_email <- read_csv("data/gss_email_2018.csv")
```


We will use the `email` variable: the number of minutes that respondents spend on email weekly. 

The goal of this exercise is to compute the 95% confidence interval for the mean time that Americans spend on email.


### 10.1

Set a seed to control the random process.

```{r}
set.seed(124534)
```

### 10.2

Compute the bootstrap distribution

```{r}
bss <- gss_email %>%
  specify(response = email) %>% 
  generate(reps = 5000, type = "bootstrap")
```



```{r glimpse-boot-dist}
glimpse(bss)
```

### 10.3

Compute and visualize the CI

```{r calc-ci}
bss_stat <- bss %>% calculate(stat="mean")
```

```{r plot-boot-dist-ggplot}
ci <- bss_stat %>%
  summarise(lower_b = quantile(stat, 0.025), 
            upper_b = quantile(stat, 0.975))
ci
```

```{r  plot-boot-dist-infer}
ggplot(data = bss_stat, mapping = aes(x = stat)) +
  geom_histogram(binwidth = 5, alpha = .5) +
  geom_vline(xintercept = ci$lower_b,
             color = "steelblue", lty = 2, size = 1) + 
  geom_vline(xintercept = ci$upper_b,
             color = "steelblue", lty = 2, size = 1) +
  labs(title = "Bootstrap distribution of the mean time spent on email",
       subtitle = "with 95% CI",
       x = "Means", y = "Count") +
  theme_minimal()

```

### 10.4 

Interpret the result



## Exercise 11

So far we have worked only with models in which the response variable is quantitative.
There are analogous models that can be used for cases in which the response variable is qualitative. One of these models is logistic regression.
We will not explain how logistic regression works, but we are going to learn how a logistic regression is fitted in R.

If our response variable `y` is qualitative and we have, for example, two predictors `x1` and `x2`, we fit a logistic regression as follows:

```{r, eval=F}
log_reg <- glm( y ~ x1 + x2, data = data, family = binomial(link='logit'))
```

We are going to use a logistic regression to try to predict whether the Titanic passengers survived or not, using as predictor variables, for example, sex, age, the class of their ticket, etc.

First install the `titanic` package and load the data:

```{r}
library(titanic)
data("titanic_train")
```

Select the variables Survived, Age, Sex, Embarked, SibSp, Parch, Fare, Pclass.
Convert to factor variables those that you consider to be factors. In addition,
add an id variable indicating the row number. Save this in a new data frame called `titanic`.

**Hint:** use `?titanic_train` to see what each variable means.

### Solution

```{r}
titanic <- titanic_train %>% select(Survived, Age, Sex, Embarked, SibSp, 
                                    Parch, Fare, Pclass) %>% 
  mutate(Survived=as.factor(Survived), Sex=as.factor(Sex), 
         Embarked=as.factor(Embarked), Pclass = as.factor(Pclass)) %>% 
  mutate(row_num = row_number())

```

## Exercise 12

Create two datasets called `train` and `test`.
The first must contain observations (rows) selected at random from the titanic
data frame from the previous exercise. Its number of rows must be 70% of the number of rows of titanic. The second dataset must contain all the observations that are in the titanic dataset but are not in `train`.

**Hint:** use the `sample_frac()` function

### Solution

```{r}
train <- titanic %>% sample_frac(size=0.7)
test <- titanic %>% anti_join(train, by="row_num")
```

## Exercise 13

Fit a logistic regression model using the `train` data where the response
variable is `Survived` and the predictor variable is `Sex`.

### Solution

```{r}
log_reg <- glm( Survived ~ Sex, data = train, family = binomial(link='logit'))
```


## Exercise 14

The prediction of a logistic regression is a probability corresponding to the probability that the response variable takes value 1 (in our case that the corresponding passenger survives) according to the model.

Compute the predictions of the previous model on the `test` data. You will need to add type="response" to the add_predictions function. Build a new variable that takes value 1 if the predicted probability is greater than or equal to 0.5 and zero otherwise.

Finally, compute the percentage of correct predictions by comparing the newly created variable and the Survived variable.

### Solution

```{r}
test %>% add_predictions(log_reg, type = "response") %>% mutate(Survived_pred = ifelse(pred > 0.5, 1, 0)) %>% 
  summarise(mean(Survived == Survived_pred))
```


## Exercise 15

Try to improve the previous model by adding more variables. Which variable do you think has the most predictive power?

### Solution

