---
title: "Lab 1 - Programming in R"
output:
  html_document:
    css: ./ds_labs.css
    toc: yes
    toc_float: yes
    fig_caption: no
    seal : False
    includes:
      after_body: insert-logo.html
  pdf_document:
    toc: yes
---


<div class="title-logo"></div>

```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=8, fig.height=6) 
```


## Exercise 1

Write a function that returns $k$ numbers, where each number is
the sum of $n$ numbers drawn at random between 1 and 100 (with replacement), divided by $n$.

Plot a histogram of the results for $k=10000$ with $n=1$, $n=2$, $n=10$ and $n=100$.
Hint: you can use the `hist()` function.

What do you observe?

### Solution

We define the function

```{r}
gen <- function(k=10, n=5){
  
  result <- c()
  for(i in 1:k){
    nums <- sample(1:100, size=n, replace=TRUE)
    result <- c( result, sum(nums)/n )
  }

  return(result)
}

```

We plot the histograms

* $n=1$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen(k=10000, n=1), breaks = 50 )
```

* $n=2$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen(k=10000, n=2), breaks = 101 )
```

* $n=10$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen(k=10000, n=10), breaks = 50 )
```

* $n=100$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen(k=10000, n=100), breaks = 50 )
```


The distribution approaches a **normal distribution**.


## Exercise 2

Use the help page of the `replicate()` function to learn how it works.
Solve Exercise 1 without using a loop (if you used one).

### Solution

We define the function, this time without loops

```{r}
aux <- function(n=5){

  nums <- sample(1:100, size=n, replace=TRUE)
  return( sum(nums)/n )
}

gen2 <- function(k=10, n=5){
  return( replicate(k, aux(n)) )
}
```


* $n=1$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen2(k=10000, n=1), breaks = 50 )
```

* $n=2$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen2(k=10000, n=2), breaks = 50 )
```

* $n=10$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen2(k=10000, n=10), breaks = 50 )
```

* $n=100$

```{r,fig.align='center',out.extra='angle=90'}
hist( gen2(k=10000, n=100), breaks = 50 )
```


## Exercise 3

R's `mtcars` dataset contains information extracted from the *1974 Motor Trend US magazine* about 10 aspects of design and performance for 32 vehicles. 
Using this dataset, compute 

**Question 1.** The difference between the average fuel consumption of automatic-transmission vehicles and that of manual-transmission ones.

**Question 2.** The same difference, but only for vehicles weighing more than 3000 lbs.

Hint: the `?mtcars` help page may be useful

### Solution

**Question 1.**
```{r}
mean( mtcars[mtcars$am == 0, "mpg"] ) - mean( mtcars[mtcars$am == 1, "mpg"] )
```

**Question 2.**
```{r}
mean( mtcars[mtcars$am == 0 & mtcars$wt > 3, "mpg"] ) - mean( mtcars[mtcars$am == 1 & mtcars$wt > 3, "mpg"] )
```

## Exercise 4

Using a `for` loop and an `if`, add up the even numbers between 1 and 100 (both included).

### Solution

```{r}
sum <- 0
for(i in 1:100){
  if(i%%2 == 0){
    sum <- sum + i
  }
}
sum
```

Without a loop we could do:
```{r}
sum(seq(2,101,by=2))
```

## Exercise 5

Compute the expected gain of the roulette strategy that consists of always betting 10 euros on red.

To do so:

  - Create a function called `spin()` that returns a number drawn at random between 0 and 36 (both included).
  - Create another function called `compute_gain()` with an argument `q` for the amount of money bet, with a default value of 10. This function has to generate one spin of the roulette wheel and, depending on the outcome, compute the gain and return it. Remember that
  
    * The numbers (32, 19, 21, 25, 34, 27, 36, 30, 23, 5, 16, 1, 14, 9, 18, 7, 12, 3) are red.
    * The numbers (15, 4, 2, 17, 6, 13, 11, 8, 10, 24, 33, 20, 31, 22, 29, 28, 35, 26) are black.
    * Zero has no colour.
    
  - Finally, create a loop that simulates `K = 10000` rounds of roulette, storing the gain generated in each one, thus building a vector of length $K$. Call this vector `winnings`. Compute its mean. What do you observe?
  
### Solution

We first create the `spin()` function
```{r}
spin <- function(){
  sample(0:36, size = 1)
}
```

And the `compute_gain()` function
```{r}
red   <- c(32, 19, 21, 25, 34, 27, 36, 30, 23, 5, 16, 1, 14, 9, 18, 7, 12, 3)
black <- c(2,4,6,8,10,20,22,24,26,28,11,13,15,17,29,31,33,35)

compute_gain <- function(q=10){
  result <- spin()
  if(result %in% red){
    return(q)
  }
  else{
    return(-q)
  }
}
```

Finally we write the loop
```{r}
K=10000
winnings <- c()
for(i in 1:K){
  winnings <- c(winnings, compute_gain(1))
}
```

We can represent this graphically
```{r}
library(ggplot2)

results = data.frame(Round = 1:K, Cumulative_Winnings = cumsum(winnings) )

p <- ggplot(results, aes(x=Round, y=Cumulative_Winnings)) +  
  geom_line() + theme_minimal()

p
```


## Exercise 8

Write R code to estimate the value of $\pi$ using the Monte Carlo method. The method consists of generating random points in the unit square and counting how many of them fall inside a quarter circle inscribed in that square.

1. Generate $N$ random points inside a square of side 1, that is, the points $(x, y)$ must have coordinates $x$ and $y$ between 0 and 1. You can use the `runif()` function for this.

2. Compute the number of points that fall inside the quarter circle. 

3. Use the number computed in the previous step and the total number of points generated to estimate the value of $\pi$.

What do you observe as you increase the number of points generated?


### Solution

```{r}
N <- 10000
x <- runif(N)
y <- runif(N)
# Compute how many points fall inside the quarter circle
points_inside <- sum(x^2 + y^2 <= 1)

# Estimate the value of pi: area of the quarter circle (pi/4) times the area of the square (1)
pi_estimate <- (points_inside / N) * 4

```



