---
title: "Lab 2.2 - Data Transformation"
output:
  html_document:
    css: ./ds_labs.css
    toc: yes
    toc_float: yes
    fig_caption: no
    seal : False
    includes:
      after_body: insert-logo.html
  pdf_document:
    toc: yes
---


<div class="title-logo"></div>

```{r packages_setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
```


In this lab we will work with data from a survey on the habits of smokers in the United Kingdom. The data come from the National STEM Centre, Large Datasets from stats4schools, [https://www.stem.org.uk/resources/elibrary/resource/28452/large-datasets-stats4schools](https://www.stem.org.uk/resources/elibrary/resource/28452/large-datasets-stats4schools).

The data are in the data folder of the repository. You can load them using

```{r load-data, message=FALSE, eval=TRUE}
library(tidyverse)
smoking <- read_csv("data/smoking.csv")
```
The variables collected are (the levels they take appear in parentheses):

* `gender`: Gender

* `age`: Age

*  `marital_status`: Marital status (Divorced, Married, Separated, Single and Widowed)

* `highest_qualification`: Highest qualification (A Levels, Degree, GCSE/CSE, GCSE/O Level, Higher/Sub Degree, No Qualification, ONC/BTEC and Other/Sub Degree)

* `nationality`: Nationality (British, English, Irish, Scottish, Welsh, Other, Refused and Unknown)

* `ethnicity` : Ethnicity  (Asian, Black, Chinese, Mixed, White and Refused Unknown)

* `gross_income`: Gross income (Under 2,600, 2,600 to 5,200, 5,200 to 10,400, 10,400 to 15,600, 15,600 to 20,800, 20,800 to 28,600, 28,600 to 36,400, Above 36,400, Refused and Unknown)

* `region`: Region (London, Midlands & East Anglia, Scotland, South East, South West, The North and Wales)

* `smoke` : Smoker (No and Yes)

* `amt_weekends`: Number of cigarettes per day on weekends

* `amt_weekdays`:  Number of cigarettes per day on weekdays

* `type`: Type of cigarettes (Packets, Hand-Rolled, Both/Mainly Packets and Both/Mainly Hand-Rolled)



## Exercise 1

How many observations are there in the dataset for each gender? What proportion of women are smokers? And of men?

### Solution

```{r}
smoking %>% group_by(gender, smoke) %>% summarise(n = n()) %>% 
  mutate(proportion = n/sum(n)) 
```
24.2% of women are smokers.

## Exercise 2

Select the `amt_weekends` variable, sort it in decreasing order and show the first 10 values.
Do the same with  `amt_weekdays`; what do you observe?

### Solution

```{r}
smoking %>% select(amt_weekends) %>%
  arrange(desc(amt_weekends)) %>% slice(1:10)
```
```{r}
smoking %>% select(amt_weekdays) %>%
  arrange(desc(amt_weekdays)) %>% slice(1:10)
```


## Exercise 3

What unique values does the `gross_income` variable take?

Replace the Unknown values with `NA` and compute the unique values of the variable again.

### Solution

```{r}
smoking %>% distinct(gross_income)
#
smoking2 <- smoking %>% mutate(gross_income = ifelse(gross_income == "Unknown", NA, gross_income))

smoking2 %>% distinct(gross_income)
```


## Exercise 4 

Show the last 10 rows of the dataset.

### Solution

```{r}
smoking %>% slice_tail(n = 10)
```

```{r}
smoking %>% slice( (nrow(smoking) - 9) : nrow(smoking))
```

## Exercise 5

What is the mean number of cigarettes per weekday smoked by smokers under 30? 
And the mean number of cigarettes per weekday smoked by under-30s in general?

**Note**: If you see `NA`, determine  the possible origin of those missing values and how to handle them.

### Solution

```{r}
smoking %>% filter(age<30, smoke=="Yes") %>%  summarise(mean = mean(amt_weekends))
```


```{r}
smoking %>% filter(age<30, smoke == "Yes") %>%   summarise(mean_we = mean(amt_weekends), mean_wd = mean(amt_weekdays))
```


## Exercise 6

Compute the mean, median, and standard deviation of cigarettes/day on weekends and on weekdays by age groups less than or equal to 20, 20-30, ..., >90. 
Summarize the conclusions drawn from the results observed.

*Note*: Create a new variable called `decade` using the `case_when` function.
The levels of this variable must be
`"<=20","20-30","30-40", "40-50", "50-60", "60-70", "70-80", "80-90", ">90"`.


### Solution

```{r}
smoking_new <- smoking %>% mutate(decade=case_when(age<=20 ~"<=20",
                    age>20 & age<=30 ~ "20-30",
                    age>30 & age<=40 ~ "30-40",
                    age>40 & age<=50 ~ "40-50",
                    age>50 & age<=60 ~ "50-60",
                    age>60 & age<=70 ~ "60-70",
                    age>70 & age<=80 ~ "70-80",
                    age>80 & age<=90 ~ "80-90",
                    TRUE ~ ">90"))

smoking_new %>% filter(smoke == "Yes") %>% 
  group_by(decade) %>%
  summarise(mean_we = mean(amt_weekends), 
            mean_wd = mean(amt_weekdays),
            median_we = median(amt_weekends),
            median_wd = median(amt_weekdays),
            sd_we = sd(amt_weekends),
            sd_wd = sd(amt_weekdays) )
```


## Exercise 7

Sort the different ethnicities in decreasing order of the proportion of smokers in the sample. Exclude those observations with unknown or undeclared ethnicity.

### Solution

```{r}
smoking %>% filter(! (ethnicity %in% c("Unknown", "Refused")) ) %>%
  group_by(ethnicity, smoke) %>% 
  summarize(n = n()) %>%
  mutate(prop_smokers = n/sum(n)) %>%
  filter(smoke == "Yes") %>%
  arrange(desc(prop_smokers))
```

## Exercise 8

Reproduce the following plot. Draw some conclusions.

### Solution

```{r}
smoking %>% mutate(decade=case_when(age<=20 ~"0-20",
                    age>20 & age<=30 ~ "20-30",
                    age>30 & age<=40 ~ "30-40",
                    age>40 & age<=50 ~ "40-50",
                    age>50 & age<=60 ~ "50-60",
                    age>60 & age<=70 ~ "60-70",
                    age>70 & age<=80 ~ "70-80",
                    age>80 & age<=90 ~ "80-90",
                    TRUE ~ "90-")) %>% 
  filter(smoke == "Yes") %>% 
  group_by(decade, gender) %>%
  summarise(median_we = median(amt_weekdays)) %>% 
  ggplot(aes(x=decade, y=median_we)) + geom_bar(stat="identity", fill="#D55E00") + 
  facet_wrap(gender~.) + theme_minimal() + labs(x="Decade", y="Median Cigarettes/Weekday")
```

## Exercise 9

Which types of tobacco are most popular in the different gross income brackets? And in the different age brackets?

*Note*: use the decade variable created earlier

### Solution

```{r}
smoking %>% filter(smoke == "Yes") %>% group_by(gross_income, type) %>%
  summarise(n = n()) %>%
  mutate(prop_type = n/sum(n))
```


```{r}
smoking_new %>% filter(smoke == "Yes") %>% group_by(decade, type) %>%
  summarise(n = n()) %>%
  mutate(prop_type = n/sum(n))
```

## Exercise 10

What is the difference between these two pieces of code?

```{r, eval=F}
smoking %>% count(gender, smoke) %>% mutate(prop=n/sum(n))
```

```{r, eval=F}
smoking %>% group_by(gender, smoke) %>% summarise(n=n()) %>%
  mutate(prop=n/sum(n))
```

### Solution

In the first there is no grouping, and therefore the percentage shown is over the total.
In the second, because `group_by` is used, the percentage shown is over the individuals 
in each group.

## Exercise 11

Among people who are not married, how many cigarettes/weekday are smoked on average by level of education?


### Solution


```{r}
smoking %>% filter(smoke == "Yes", marital_status != "Married") %>% 
  group_by(highest_qualification) %>%
  summarise(mean = mean(amt_weekdays))
```


## Exercise 12

On many occasions, in the presence of missing values, it is useful to replace them
with a value. This process is known as imputation.
One of the most common techniques for imputing categorical variables consists
of replacing the missing values with the most likely value in a certain 
segment of the population (the mode).

For example, we can replace the unknown or undeclared values of the `ethnicity` variable
with the most frequent ethnicity in the data, and assume that the error rate will be small.
A perhaps smarter strategy is to replace the unknown or undeclared values of the `ethnicity` variable with the most prevalent ethnicity in the corresponding nation. In this exercise,
you have to do the latter. To do so, use the following function
to compute the mode.

```{r}
calc_mode <- function(x){
  
  x <- x[!x %in% c("Refused", "Unkown")]
  distinct_values <- unique(x)
  distinct_tabulate <- tabulate(match(x, distinct_values))
  distinct_values[which.max(distinct_tabulate)]
  
}
```

*Note*: Exclude the observations with unknown or undeclared nationality.
What happens if you include them?

### Solution

```{r}
smoking %>% filter(!nationality %in%  c("Unknown", "Refused")) %>% 
  group_by(nationality) %>% 
  mutate(ethnicity_nona = if_else(ethnicity %in% c("Unknown", "Refused"), 
                         calc_mode(ethnicity), 
                         ethnicity))  %>% 
  filter(ethnicity %in% c("Unknown", "Refused")) %>%
  select(nationality, ethnicity, ethnicity_nona)
```


